---
title: "Final Project"
output:
  html_document:
    df_print: paged
---

```{r message=FALSE}
library(atus)
library(dplyr)
library(rpart)
library(tidyr)
library(ggplot2)
library(RColorBrewer)
library(rpart)
library(rpart.plot)
library(splines)
```

# Refined Goals of Analysis

Our group is using the atus database to investigate whether various demographic variables correlate to family income. Specifically, we will be looking at region, state, educational attainment, home ownership (rent vs. own), household size, race, and time usage. After visualizing the data, we will test the correlations between each of these demographic variables and income using linear regression or other related statistical analyses. Using the results from these statistical analyses, we will determine relationships between the demographic variables and how they collectively impact family income and how income potentially affects time usage. This information could be useful to better inform policy regarding issues relevant to specific demographics.



# Exploring the raltionship between home type and family income

```{r}
library(tidyr)
library(dplyr)
library(atus)
data("atuscps")
tibble1 <- as_tibble(atuscps) %>%
  na.omit()
```

```{r}
library(ggplot2)
library(RColorBrewer)
tibble1 <- as_tibble(atuscps) %>%
  na.omit()
plotcps <- ggplot(data=tibble1) +
  geom_bar(mapping=aes(x=famincome, fill=home_type)) +
  xlab("Family Income") +
  ylab("Count") +
  scale_color_brewer("Set1")
plotcps + theme(axis.text.x = element_text(angle = 27, size = 6)) 
```

```{r}
library(ggplot2)
library(RColorBrewer)
plotdps <- ggplot(data=tibble1) +
  geom_bar(mapping=aes(x=home_type, fill=famincome)) + 
  xlab("Home Type") +
  ylab("Count") +
  scale_color_brewer("Set1")
plotdps + theme(axis.text.x = element_text(angle = 27, size = 6))
```

## VISUALIZATION ANALYSIS:
Color-coded simple stacked bar charts, using ggplot2, were chosen to visualize the data because this method is useful in comparing the total amounts across each segmented group. Additionally, stacked bar charts have the advantages of being able to easily view the data and identify potential outliers.

When considering the relationship between the type of home (own, rent, occupy without rent) and family income, the two stacked bar charts visualize the following. No matter what family income range one falls under, there seems to be worth and demand in owning a house. This means even within a humble, lower family income range, there are still home-owners present. This offers the potentially interesting finding that people of lower socioeconomic classes are still trying to chase “the American Dream” or stability, despite renting being a more feasible option for their financial situation. It must be said of course, that this idea is not generalizable to that socioeconomic class as a whole because as both stacked bar charts suggest visually, for lower socioeconomic classes, the renter population still remains more popular than the homeowner population. On the contrary, we see that in the middle to higher income brackets, there are certainly a much larger count of homeowners compared to its opposing category, renters. 

```{r}
library(atus)
chart = table(atuscps$famincome, atuscps$home_type) 
chart
chisq.test(chart)
```

## RELATIONSHIP SIGNIFICANCE ANALYSIS (Pearson's Chi-Squared Test):
As the p-value (<2.2e-16) is less than the .05 significance level, and the chi-squared value is notably large (26953), we reject the null hypothesis and conclude that the two variables, family income and home type (own, rent, occupy without rent), are in fact dependent and have a significant relationship. 

# Exploring the relationship between household size and Income

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(atus)
atus2 <- cbind(atuscps, atusresp) 
atusresp$hh_size <- as.character(as.numeric(atusresp$hh_size))
atus3 <- atus2 %>%
  select(famincome, hh_size)
atus4 <- as_tibble(atus3)
chart1 <- ggplot(data=atus4) +
  geom_bar(mapping=aes(x=famincome, fill=hh_size)) + 
  xlab("Family Income") +
  ylab("Count") +
  scale_color_brewer("Set1") 
chart1 + theme(axis.text.x = element_text(angle = 27, size = 6))
```

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(atus)
atus2 <- cbind(atuscps, atusresp) 
atusresp$hh_size <- as.character(as.numeric(atusresp$hh_size))
atus3 <- atus2 %>%
  select(famincome, hh_size)
atus4 <- as_tibble(atus3)
chart14 <- ggplot(data=atus4) +
  geom_bar(mapping=aes(x=hh_size, fill=famincome)) + 
  xlab("Household Size") +
  ylab("Count") +
  scale_color_brewer("Set1") 
chart14 + theme(axis.text.x = element_text(angle = 27, size = 6))
```

## VISUALIZATION ANALYSIS:
The stacked bar charts depicting the relationship between household size (1-16) and family income suggest that there is a varied and potentially similar distribution of household size across each family income range. Further, in taking a closer look at the distribution of family income across household sizes, it is clear that there is a varied distribution, with no income group particularly prominent in each household size. Hence, we must use statistical analyses to see if household size and family income have a significant relationship.

```{r}
library(atus)
chart2 = table(atuscps$famincome, atusresp$hh_size) 
chart2
chisq.test(chart2)
```

## RELATIONSHIP SIGNFICANCE ANALYSIS (Pearson's Chi-Squared Test):
As the p-value (<2.2e-16) is less than the .05 significance level, and the chi-squared value is notably large (21089), we reject the null hypothesis and conclude that the two variables, family income and household size (1-16), are in fact dependent and have a significant relationship. 

\newpage

# Impact of demographic variables on time usage

## Code Setup

```{r setup}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, fig.width = 9, fig.height = 5.5, 
                      out.width = "90%", fig.align = "center", cache = T) 
```


## Data Manipulation

The ATUS data is grouped into three separate datasets--atusact, atuscps, and atusresp--that contain information on peoples' time usage, demographics, and personal information respectively. We will begin by bringing in the atusact dataset and ungrouping it so that we can start manipulating it as we wish.

```{r}
time_data = ungroup(atusact)
glimpse(time_data)
```

In its original state, atusact contains tiercodes with three, two digit pairs. In this combined format the tiercodes provide very detailed information into the exact activities the individual is performing for whatever length of time; however, what we really want to examine is the overall time various people spend on certain categories of tasks that encompass a broad range of related activities. Therefore, what we need to do is isolate the two digit pairings in the tiercodes to gain a better view of this. As you can see, this simply requires some arithmetic to separate out the various tiers.  

```{r}
time_data2 = time_data %>% 
  group_by(tiercode) %>%
  summarise(tucaseid = tucaseid,
            tier1_2Code = tiercode %/% 100,
            tier1 = tiercode %/% 10000,
            tier2 = (tiercode %% 10000) %/% 100, 
            tier3 = (tiercode %% 10000) %% 100,
            dur = dur) %>%
  inner_join(atuscps) %>% 
  select(tiercode, tier1, dur, region, state, famincome, edu) %>%
  drop_na()
time_data2
```

Additionally, as shown above, we want to join the atusact dataset with atuscps so that we can begin analyzing the relationships between peoples' time usage and various demographics.

#### Family Income  
&nbsp;

First, we will look into the impact of family income on time usage.

```{r}
time_data3 = time_data2 %>%
  group_by(tier1, famincome) %>%
  mutate(TDur = sum(dur)) %>% 
  select(famincome, tier1, TDur) %>% 
  distinct() %>% 
  arrange(tier1, TDur)
time_data3
```

#### Geographic Region  
&nbsp;

Next, we will examine the impact of region on time usage

```{r}
time_data4 = time_data2 %>%
  group_by(tier1, region) %>%
  mutate(RDur = sum(dur)) %>% 
  select(region, tier1, RDur) %>% 
  distinct() %>% 
  arrange(tier1, RDur)
time_data4
```

#### Educational Attainment  
&nbsp;

And finally, we will look into the impact of educational attainment on time usage

```{r}
time_data5 = time_data2 %>%
  group_by(tier1, edu) %>%
  mutate(EDur = sum(dur)) %>% 
  select(edu, tier1, EDur) %>% 
  distinct() %>% 
  arrange(tier1, EDur)
time_data5
```

## Generating Graphics

For each of our demographic variables (famincome, region, and edu) we will create a stacked bar graph that displays the percentage of the time that each subgroup of the demographic spends on certain types of activities (i.e. Personal Care)

```{r}
time_data3$tier1 = as.factor(time_data3$tier1)
colourCount = length(unique(time_data3$tier1))
getPalette = colorRampPalette(brewer.pal(12, "Paired"))
ggplot(data = time_data3, mapping = aes(x = famincome, y = TDur, 
                                                      fill = tier1)) +
  labs(title = "The Impact of Income on Time Usage Distribution", 
       subtitle = "Data from the 2003-2016 American Time Usage Survey", 
       y = "Time Usage (% of Total)", x = "Family Income") +
  scale_fill_brewer(palette = "Set1") +
  geom_col(position = "fill") + 
  scale_y_continuous(breaks = seq(0.0, 1.0, 0.1), labels = function(x){
    paste0(x * 100, '%')}) +
  theme_bw()  +
  scale_fill_manual(name = "Category", labels = c("Personal Care", "Household",
                    "Caring for HH Members", "Caring for NonHH Members", "Work",
                    "Education", "Shopping", "Professional Serv", 
                    "Household Serv", "Gov Serv & Civic Obligations", 
                    "Eating & Drinking", "Socializing & Leisure", 
                    "Exercise & Recreation", "Religious & Spiritual",
                    "Volunteering", "Telephone Calls", "Traveling"), 
                    values = getPalette(colourCount)) + 
  coord_flip() 
```
Upon cursory inspection, it appears that as family income increases, people tend to spend less time on Personal Care (includes sleeping) and Socializing & Leisure related activities, while spending more time Traveling, on Exercise & Recreation, and on Work related activities.

```{r}
time_data4$tier1 = as.factor(time_data4$tier1)
ggplot(data = time_data4, mapping = aes(x = region, y = RDur, fill = tier1)) +
  labs(title = "The Impact of Region on Time Usage Distribution", 
       subtitle = "Data from the 2003-2016 American Time Usage Survey", 
       y = "Time Usage (% of Total)", x = "Region") +
  geom_col(position = "fill") +
  scale_y_continuous(breaks = seq(0.0, 1.0, 0.1), labels = function(x){
    paste0(x * 100, '%')}) +
  theme_bw()  +
  scale_fill_manual(name = "Category", labels = c("Personal Care", "Household",
                    "Caring for HH Members", "Caring for NonHH Members", "Work",
                    "Education", "Shopping", "Professional Serv",
                    "Household Serv", "Gov Serv & Civic Obligations",
                    "Eating & Drinking", "Socializing & Leisure",
                    "Exercise & Recreation", "Religious & Spiritual",
                    "Volunteering", "Telephone Calls", "Traveling"),
                    values = getPalette(colourCount)) +
  coord_flip()
```
Looking at the resulting bar graph, it appears that region has little impact on time usage as the percentages of time spend on different types of activities across various regions of the United States line up almost perfectly. Although there are some slight deviations, for the most part it seems that region does not influence time usage significantly. 

```{r}
time_data5$tier1 = as.factor(time_data5$tier1)
ggplot(data = time_data5, mapping = aes(x = edu, y = EDur, fill = tier1)) +
  labs(title = "The Impact of Educational Attainment on Time Usage Distribution",
       subtitle = "Data from the 2003-2016 American Time Usage Survey", 
       y = "Time Usage (% of Total)", x = "Educational Attainment") +
  geom_col(position = "fill") + 
  scale_y_continuous(breaks = seq(0.0, 1.0, 0.1), labels = function(x){
    paste0(x * 100, '%')}) +
  theme_bw()  +
  scale_fill_manual(name = "Category", labels = c("Personal Care", "Household",
                    "Caring for HH Members", "Caring for NonHH Members", "Work",
                    "Education", "Shopping", "Professional Serv",
                    "Household Serv", "Gov Serv & Civic Obligations",
                    "Eating & Drinking", "Socializing & Leisure",
                    "Exercise & Recreation", "Religious & Spiritual", 
                    "Volunteering", "Telephone Calls", "Traveling"), 
                    values = getPalette(colourCount)) +
  coord_flip()
```
As one might expect, the trends in the resulting bar graph seem to reflect those that appear in the graph for family income. It is well known that there is a strong correlation between these two variables so this is not that surprising. Most notable among the differences in the visual is that it appears that as educational attainment increases, people spend more time on Work and less time on Socializing & Leisure as well as on Personal Care. We saw these same exact trends in the graph for family income if you recall the result of the earlier code chunk.

## Analyzing Relationship Strength/Correlation

Now, we will fit and test various models on the datasets we have created to determine whether these demographics truly have a significant impact on the time usage of people in the ATUS. We will try a number of models and examine the results.

#### Linear Regression
&nbsp;

We will start out by conducting some basic tests for linear regression.

```{r}
mod1 = lm(TDur ~ famincome, data = time_data3) 
summary(mod1)
```
Somewhat unexpectedly based on our observations of the bar graph, it appears that family income is not as influential as we thought on peoples' time usage. While the higher income brackets do show relatively low p-values, the only factor level that produces a p-value less than .05 is 'famincome75000-99999'.

```{r}
mod2 = lm(RDur ~ region, data = time_data4)
summary(mod2)
```
As we expected from the graph we generated, region has little impact on time usage. None of the p-values for the regions are below .05.

```{r}
mod3 = lm(EDur ~ edu, data = time_data5)
summary(mod3)
```
Again, somewhat surprising results based on the appearance of our graph. The p-values for the various educational attainments are even worse than those for family income and none possess a value below .05.

#### Multi-variable Linear Regression
&nbsp;

Lets try grouping the dataset observations by tier, famincome, region, and edu to see if we find anything different. 

```{r}
time_data6 = time_data2 %>% 
  group_by(tier1, famincome, region, edu) %>% 
  mutate(Dur = sum(dur)) %>% 
  select(tier1, famincome, region, edu, Dur) %>%
  distinct()
time_data6
```

```{r}
mod4 = lm(Dur ~ famincome + region + edu, data = time_data6)
summary(mod4)
```
The results of this tests are still rather confusing as the p-values don't line up with the results from earlier tests. The results show both educational attainment and region as having significant impact on time usage despite the results of the earlier tests. Based on this test the most significant predictors are certain levels of each factor. These levels include edudoctoral degree, eduprof degree, regionsouth, and famincome 75000-99999.

## Conclusion

Overall, based on the results of the tests and the graphics generated, I think that it is difficult to reach a decisive decision on the impact of the demographic variables Family Income, Region, and Educational Attainment on Time Usage. However, I would say that it seems these three variables do not have that significant of an impact on Time Usage. If I were to rank the impact of the variables it would likely go Family Income, Educational Attainment, and then finally Region.

\newpage

# Investigating factors which influence family income

## Here, we investigated the relationship between various demographic variables and income. 
```{r warning=FALSE}
# Find average family income by educational attainment
avg_income_by_edu <- atuscps 
avg_income_by_edu <- avg_income_by_edu %>%
  separate(famincome, into=c('income_low','income_high'), sep='-',convert=TRUE)%>%
  mutate(income_low = as.integer(income_low)) %>%
  drop_na() %>%
  mutate(fam_income_mid = (income_high+income_low)/2) %>%
  group_by(education=edu) %>%
  summarise(N = n(),
            avg_income = mean(fam_income_mid)) %>%
  arrange(avg_income)
avg_income_by_edu
```
```{r}
#Visualize results
ggplot(data=avg_income_by_edu, aes(x=education, y=avg_income)) +
  geom_bar(stat='identity',  fill = brewer.pal(n=8,name = "Set1")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Investigating income by education")
```

Here we find the expected result of an apparent trend in income as educational attainment increases. 

```{r warning=FALSE}
# Investigate by region
# Find average family income by educational attainment
avg_income_by_edu_region <- atuscps 
avg_income_by_edu_region <-avg_income_by_edu_region %>%
  separate(famincome, into=c('income_low','income_high'), sep='-',convert=TRUE)%>%
  mutate(income_low = as.integer(income_low)) %>%
  drop_na() %>%
  mutate(fam_income_mid = (income_high+income_low)/2) %>%
  group_by(region,edu) %>%
  summarise(N = n(),
            avg_income = mean(fam_income_mid)) %>%
  arrange(avg_income)
avg_income_by_edu_region %>% head(10)
```

```{r}
# Compare regions by income
region_df <- avg_income_by_edu_region %>%
  group_by(region) %>%
  summarise(N=n(),
            avg_income = mean(avg_income)) %>%
  arrange(avg_income)
ggplot(region_df, aes(x=region, y=avg_income, label=avg_income)) +
  geom_bar(stat='identity', fill = brewer.pal(n=4,name = "BuPu")) +
  geom_text(aes(label = signif(avg_income, digits = 3)), nudge_y = -2000, color='DarkBlue') + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ggtitle("Investigating income by region")
```

Here the results are less clear, although it would appear that the south has relatively low average income. 

```{r}
# Compare income by region and edu
ggplot(avg_income_by_edu_region, aes(x=region, y=avg_income)) +
    geom_col(position='stack') +
    scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    geom_text(aes(label = signif(avg_income, digits = 3)), angle = 90, 
              nudge_y = -10000, color = 'white') +
    facet_grid(~ edu) +
    scale_fill_brewer(palette="Set1") +
    ggtitle("Investigating income by education and region")
```
Here we see specific education levels may play a mitigating role in the observed disparity by region. 

```{r warning=FALSE}
avg_income_by_race <- atuscps 
avg_income_by_race <-avg_income_by_race %>%
  separate(famincome, into=c('income_low','income_high'), sep='-',convert=TRUE)%>%
  mutate(income_low = as.integer(income_low)) %>%
  drop_na() %>%
  mutate(fam_income_mid = (income_high+income_low)/2) %>%
  group_by(race,region) %>%
  summarise(N = n(),
            avg_income = mean(fam_income_mid)) %>%
  arrange(avg_income)
avg_income_by_race %>% head(10)
```

```{r}
ggplot(avg_income_by_race, aes(x=region, y=avg_income)) +
    geom_bar(position='stack', stat='identity') +
    scale_fill_manual(palette = "Set1") +
    scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    geom_text(aes(label = signif(avg_income, digits = 3)), angle = 90, 
              nudge_y = -5500, color = 'white') +
    facet_grid(~ race) +
    ggtitle("Investigating income by race and region")
```
The results here show a more nuanced picture. The effects of region on income are mitigated by race differently. 

```{r warning=FALSE}
avg_income_by_race_edu <- atuscps 
avg_income_by_race_edu <-avg_income_by_race_edu %>%
  separate(famincome, into=c('income_low','income_high'), sep='-',convert=TRUE)%>%
  mutate(income_low = as.integer(income_low)) %>%
  drop_na() %>%
  mutate(fam_income_mid = (income_high+income_low)/2) %>%
  group_by(race,edu) %>%
  summarise(N = n(),
            avg_income = mean(fam_income_mid)) %>%
  arrange(avg_income)
avg_income_by_race_edu
```
```{r}
ggplot(avg_income_by_race_edu, aes(x=edu, y=avg_income)) +
    geom_col() +
    scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
    scale_fill_brewer(palette="Set1") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    geom_text(aes(label = signif(avg_income, digits = 3)), angle = 90, 
              nudge_y = -10000, color = 'white') +
    facet_grid(~ race) +
    ggtitle("Investigating income by race and education")
```
The generally linear appearance of the graph is consistent with the above findings excluding the "Other" categor which has an unusal 

### Analysis

```{r warning=FALSE}
# Put to the test
atuscps_mid_income <- atuscps 
atuscps_mid_income <- atuscps_mid_income %>%
  separate(famincome, into=c('income_low','income_high'), sep='-',convert=TRUE)%>%
  mutate(income_low = as.integer(income_low)) %>%
  drop_na() %>%
  mutate(fam_income_mid = (income_high+income_low)/2)
test_stats <- lm(fam_income_mid ~ edu + region + race, data=atuscps_mid_income)

summary(test_stats)
```

Here we can see the initial hypothesis was confirmed ane race, education, and region are all significant predictors of family income. 

\newpage

# Exploring the relationship bewteen Race and Income 

##Background Visualizations

### Racial Composition of U.S.
  The following bar graph represents the numbers of people of each race in the U.S. population. This provides a frame of reference for the U.S. population composition when viewing later data visualizations.

*National Visualization*
```{r}
race_plot <- ggplot(data = atuscps, aes(x = race, fill = race)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Racial Composition of the U.S.", x = NULL, y = NULL, fill = "Race") +
  labs(title = "Regional Racial Compositions in the U.S.", x = "Race", y = "Count") +
  theme(
    plot.title = element_text(hjust = 0.5)
    ) 
race_plot
```

*Regional Visualization*
  The regional racial distributions seem to be similar to the national one. From this visualization, it can be concluded that the most populated region of the U.S. is the South and the least is the Northeast.
  
```{r}
race_plot2 <- ggplot(data = atuscps, aes(x = race, fill = race)) +
  geom_bar() +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Racial Composition of the U.S.", x = NULL, y = NULL, fill = "Race") +
  labs(title = "Regional Racial Compositions in the U.S.", x = "Race", y = "Count") +
  theme(
    plot.title = element_text(hjust = 0.5)
    ) +
  facet_wrap(~ region)
race_plot2
```

### Subcategories of Race
  Other demographic variables, such as citizenship status and country of origin, are relevant to race and may act as subcategories of race.

#### Citizenship Status Across Race
*National Visualization*
  Surprisingly, there seems to be a greater percentage of black people who hold citizenship than that of white people. This may be due to greater immigration from European countries. It is clear that Asians are most likely to not hold citizenship status among the racial categories studided.
```{r}
citizenship_plot <- ggplot(data = atuscps, mapping = aes(x = race, fill = citizen)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "U.S. Citizenship Status Across Races", x = "Race", y = "Count", fill = "U.S. Citizen") +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
citizenship_plot
```

*Regional Visualization*
  Aside from Asians, every other racial group consistently has at least 87.5% citizenship in each region. In the west, a greater proportion of Asians are citizens compared to those of the three other regions.
  
```{r}
citizenship_plot2 <- ggplot(data = atuscps, mapping = aes(x = race, fill = citizen)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "U.S. Citizenship Status Across Races by Region", x = "Race", y = "Count", fill = "U.S. Citizen") +
  theme(
    plot.title = element_text(hjust = 0.5)
  ) +
  facet_wrap(~ region)
citizenship_plot2
```

#### Country of Origin
  In the `atuscps` dataset, the variable listing each respondent's country of origin is binary; repsondents were asked whether or not their country of origin is the U.S. Both the national and regional visualizations for this variable seem to correspond strongly with the U.S. citizenship status visualizations in the previous section.

*National Visualization*
```{r}
country_plot <- ggplot(data = atuscps, mapping = aes(x = race, fill = country_born)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Countries of Origin Across Races", x = "Race", y = "Count", fill = "Country of Origin") +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
country_plot
```

*Regional Visualization*
```{r}
country_plot2 <- ggplot(data = atuscps, mapping = aes(x = race, fill = country_born)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Countries of Origin Across Races by Region", x = "Race", y = "Count", fill = "Country of Origin") +
  theme(
    plot.title = element_text(hjust = 0.5)
  ) +
  facet_wrap(~ region)
country_plot2
```

### Race vs. Income
  To visualize the possible relationship between race and income, three representations of income were used : family income brackets, average family income, and the ratio of the top 1% to the bottom 99%. 

  Observations with missing values have been removed from this dataset to better visualize the income graphs
```{r}
income <- atuscps %>% 
  na.omit()
income
```

#### Family Income Brackets

*Race*
  As expected, whites and Asians are increasingly larger proportions of income brackets as they ascend. Conversely, blacks and other racial groups are decreasingly smaller proportions of income brackets as they ascend. This income disparity could be attributed to economic and social inequality.
  
```{r}
race_income <- ggplot(data = income, mapping = aes(x = famincome, fill = race)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Racial Breakdown of Each Income Bracket in the U.S.", x = "Family Income Bracket", y = "Count", fill = "Race") +
  theme(
    plot.title = element_text(hjust = 0.5)
  )
race_income
```

*Regional Differences in Income*
  To determine whether the regional income distributions differed from those of the national visualization, the `facet_wrap()` function was used. Again, the regional trends are similar to those of the national one.
  
```{r}
race_income2 <- ggplot(data = income, mapping = aes(x = famincome, fill = race)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Racial Breakdown of Each Income Bracket in the U.S. by Region", x = "Family Income Bracket", y = "Count", fill = "Race") +
  theme(
    plot.title = element_text(hjust = 0.5)
  ) +
  facet_wrap(~ region)
race_income2
```

*U.S. Citizenship*
  While the proportion of non-citizen blacks in each income bracket remains fairly consistent, the proportion of blacks with U.S. citizenship decreases as the income brackets increase. For whites, the proportion of people decreases for non-citizens and increases for citizens as the income brackets increase. The other racial groups seem to maintain the same distributions regardless of citizenship status. Although this requires further statistical analysis, these results may indicate that citizenship is a significant variable in predicting income.
  
```{r}
race_income3 <- ggplot(data = income, mapping = aes(x = famincome, fill = race)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Racial Breakdown of Each Income Bracket in the U.S. by Citizenship", x = "Family Income Bracket", y = "Count", fill = "Race") +
  theme(
    plot.title = element_text(hjust = 0.5)
  ) +
  facet_wrap(~ citizen)
race_income3
```

*Country of Origin*
  The proportion of whites born outside of the U.S. decreases as the income bracket increases, which corresponds to the trend seen for non-citizen whites in the citizenship visualization. The proportion of blacks born outside of the U.S. remains fairly consistent across each income bracket, which again reflects the citizenship distribution. Distributions for the two other racial groups are similar to those of the national ones.
  
```{r}
race_income4 <- ggplot(data = income, mapping = aes(x = famincome, fill = race)) +
  geom_bar(position = "fill") +
  scale_fill_brewer(palette = "Set1") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Racial Breakdown of Each Income Bracket in the U.S. by Country of Origin", x = "Family Income Bracket", y = "Count", fill = "Race") +
  theme(
    plot.title = element_text(hjust = 0.5)
  ) +
  facet_wrap(~ country_born)
race_income4
```

#### Average Family Income
  The dataset below calculates the average family income for each race within each region.
```{r}
avg_income_by_race <- atuscps 
avg_income_by_race <-avg_income_by_race %>%
  separate(famincome, into=c('income_low','income_high'), sep='-',convert=TRUE)%>%
  mutate(income_low = as.integer(income_low)) %>%
  drop_na() %>%
  mutate(fam_income_mid = (income_high+income_low)/2) %>%
  group_by(race) %>%
  summarise(N = n(),
            avg_income = mean(fam_income_mid)) %>%
  arrange(avg_income)
avg_income_by_race
```

*Race*
  Surprisingly, Asians have the highest average family income, followed by whites, other, then blacks. 
```{r}
avg_income_plot <- ggplot(data = avg_income_by_race, mapping = aes(x = race, y = avg_income)) +
  geom_col() +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Average U.S. Family Income by Race", x = "Race", y = "Average Family Income") +
  theme(
    plot.title = element_text(hjust = 0.5)
  ) 
avg_income_plot
```

*Regional Differences in Income*
  The average family income maintains the same distribution for each region as the general racial one in the previous visualization.
```{r}
avg_income_by_race2 <- atuscps %>%
  separate(famincome, into=c('income_low','income_high'), sep='-',convert=TRUE)%>%
  mutate(income_low = as.integer(income_low)) %>%
  drop_na() %>%
  mutate(fam_income_mid = (income_high+income_low)/2) %>%
  group_by(race, region) %>%
  summarise(N = n(),
            avg_income = mean(fam_income_mid)) %>%
  arrange(avg_income)
avg_income_by_race2
```

```{r}
avg_income_plot2 <- ggplot(data = avg_income_by_race2, mapping = aes(x = race, y = avg_income)) +
  geom_col() +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Regional Average U.S. Family Income by Race", x = "Race", y = "Average Family Income") +
  theme(
    plot.title = element_text(hjust = 0.5)
  ) +
  facet_wrap(~ region) +
  geom_text(aes(label = signif(avg_income, digits = 3)), angle = 90, 
              nudge_y = -15000, color = 'white')
avg_income_plot2
```

*U.S. Citizenship*
   For U.S. citizens, the distribution was the same as the ones for the previous two visualizations. Surprisingly, average family income was about $40,000 for all non-citizens except for Asians. Asian non-citizens had a considerably higher average family income than these other racial groups. 
```{r}
avg_income_by_race3 <- atuscps %>%
  separate(famincome, into=c('income_low','income_high'), sep='-',convert=TRUE)%>%
  mutate(income_low = as.integer(income_low)) %>%
  drop_na() %>%
  mutate(fam_income_mid = (income_high+income_low)/2) %>%
  group_by(race, citizen) %>%
  summarise(N = n(),
            avg_income = mean(fam_income_mid)) %>%
  arrange(avg_income)
avg_income_by_race3
```

  
```{r}
avg_income_plot3 <- ggplot(data = avg_income_by_race3, mapping = aes(x = race, y = avg_income)) +
  geom_col() +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Average U.S. Family Income by Citizenship Status Across Race", x = "Race", y = "Average Family Income") +
  theme(
    plot.title = element_text(hjust = 0.5)
  ) +
  facet_wrap(~ citizen) +
  geom_text(aes(label = signif(avg_income, digits = 3)), angle = 90, 
              nudge_y = -15000, color = 'white')
avg_income_plot3
```

*Country of Origin*
  The country of origin trend mirrors that of the one for citizenship. This makes sense, as respondents who reported that they were born outside of the U.S. most likely do not hold U.S. citizenship.
```{r}
avg_income_by_race4 <- atuscps %>%
  separate(famincome, into=c('income_low','income_high'), sep='-',convert=TRUE)%>%
  mutate(income_low = as.integer(income_low)) %>%
  drop_na() %>%
  mutate(fam_income_mid = (income_high+income_low)/2) %>%
  group_by(race, country_born) %>%
  summarise(N = n(),
            avg_income = mean(fam_income_mid)) %>%
  arrange(avg_income)
avg_income_by_race4
```

  
```{r}
avg_income_plot4 <- ggplot(data = avg_income_by_race4, mapping = aes(x = race, y = avg_income)) +
  geom_col() +
  scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Average U.S. Family Income by Citizenship Status Across Race", x = "Race", y = "Average Family Income") +
  theme(
    plot.title = element_text(hjust = 0.5)
  ) +
  facet_wrap(~ country_born) +
  geom_text(aes(label = signif(avg_income, digits = 3)), angle = 90, 
              nudge_y = -15000, color = 'white')
avg_income_plot4
```

## Statistical Analysis
  First, simple linear regresions were conducted to analyze the individual demographic varaibles from the study. Then, they were combined to form a multiple regression to create a more complex and realistic relationship to income.

```{r}
income2 <- atuscps %>%
  separate(famincome, into=c('income_low','income_high'), sep='-',convert=TRUE)%>%
  mutate(income_low = as.integer(income_low)) %>%
  drop_na() %>%
  mutate(fam_income_mid = (income_high+income_low)/2) 
income2
```

```{r}
mod1 <- lm(fam_income_mid ~ race, data = income2)
summary(mod1)
```

```{r}
mod2 <- lm(fam_income_mid ~ citizen, data = income2)
summary(mod2)
```

```{r}
mod3 <- lm(fam_income_mid ~ country_born, data = income2)
summary(mod3)
```

  To represent the more complex relationships between all of these demographic variables and income, I used multiple regression 
```{r}
mod4 <- lm(fam_income_mid ~ race + citizen + country_born,  data = income2)
summary(mod4)
```

## Conclusion
  Based on the results of the linear regressions conducted, all of the demographic variables studied are significant due to their small p-values (<2e-16). Unfortunately, this means that the impacts and importance of each variable cannot be distinguished since they returned the same p-values and are all concluded to be significant. In conclusion, race, citizenship, and country of origin are all significant predictors of family income.



\newpage

# Appendix

## Contributions
### Mahima 
* Exploring the relationship between household size and Income
* Exploring the relationship between home type and family income

### Zach
* Impact of demographic variables on time usage

### James
* Investigating factors which influence family income

### Lihn
* Exploring the relationship between race and income

## Part 1 Data sets
* Mahima - drug_use
* Zach - bad_drivers
* James - endorsements_2020
* Lihn - hiphop_cand_lyrics