---
title: 'STAT 1261: Final Project (Part 2)'
author: "Linh Huynh"
date: "11/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(atus)
library(dplyr)
library(tidyr)
library(scales)
library(ggplot2)
```

```{r}
atuscps
```

### Regional Differences in Mean Income
I'm having some trouble finding a way to consolidate the data meaningfully, so I'm going to hold off on this part of the data wrangling and visualization for now. I have done a bit of work for it, though.

#### Preparing Data Set
```{r}
reg_income <- atuscps %>%
  select(region, famincome) %>%
  separate(famincome, into = c("low_famincome", "high_famincome"), sep = "-") %>%
  mutate(
    low_famincome = as.integer(low_famincome), 
    high_famincome = as.integer(high_famincome)
  ) %>%
  filter(
    !low_famincome %in% NA,
    !high_famincome %in% NA
  )
reg_income
```

#### Finding Original Income Brackets
```{r}
levels(atuscps$famincome)
```

#### Bounds of Family Income
```{r}
bounds_reg_income <- reg_income %>%
  group_by(region) %>%
  summarise(
    mean_low_fincome = mean(low_famincome),
    mean_high_fincome = mean(high_famincome)
  )
bounds_reg_income
```

#### New Income Brackets
```{r}
i_brackets <- bounds_reg_income %>%
  mutate(
    mean_low_fincome = round(mean_low_fincome),
    mean_high_fincome = round(mean_high_fincome)
  ) %>%
  unite(income_bracket, mean_low_fincome, mean_high_fincome, sep = "-")
i_brackets
```


#### Mapping

##### Lower Bound of Family Income

##### Higher Bound of Family Income


### State Differences in Mean Income



### Analyzing Race Demographics

#### Racial Breakdown
```{r}
num_resp <- nrow(atuscps); num_resp
```

```{r}
race <- atuscps %>%
  select(race) %>%
  group_by(race) %>%
  summarise(
    num_race = n()
  ) %>%
  mutate(percent_race = percent((num_race/num_resp))) 
race
```

```{r}
race_plot <- ggplot(data = atuscps, aes(x = factor(1), fill = race)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y")
race_plot
```

#### Citizenship Status Across Races
```{r}
citizenship_plot <- ggplot(data = atuscps, mapping = aes(x = race, fill = citizen)) +
  geom_bar(position = "fill")
citizenship_plot
```
Note: Go back and reorder ethnicities?

#### Country of Origin Across Races
```{r}
country_plot <- ggplot(data = atuscps, mapping = aes(x = race, fill = country_born)) +
  geom_bar(position = "fill")
country_plot
```

#### Family Income Based on Race
```{r}
race_income <- ggplot(data = atuscps, mapping = aes(x = famincome, fill = race)) +
  geom_bar(position = "fill")
race_income
```


#### Regional and State Differences in Mean Income for Minorities
Still need to complete this