---
title: "Final Project (Pt.2)"
author: "Linh Huynh"
date: "11/22/2020"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(atus)
library(tidyr)
library(dplyr)
library(scales)
library(ggplot2)
library(RColorBrewer)
```

# Race, Citizenship, and Country of Origin Background Info

## Racial Composition of U.S.
```{r}
race_color <- brewer.pal(4, "Set1"); race_color
race_plot <- ggplot(data = atuscps, aes(x = factor(1), fill = race)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y")
race_plot
```

## Citizenship Status Across Race
```{r}
citizenship_plot <- ggplot(data = atuscps, mapping = aes(x = race, fill = citizen)) +
  geom_bar(position = "fill")
citizenship_plot
```

By Region:
```{r}
citizenship_plot2 <- ggplot(data = atuscps, mapping = aes(x = race, fill = citizen)) +
  geom_bar(position = "fill") +
  facet_wrap(~ region)
citizenship_plot2
```

By State:
```{r}
citizenship_plot3 <- ggplot(data = atuscps, mapping = aes(x = race, fill = citizen)) +
  geom_bar(position = "fill") +
  facet_wrap(~ state) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
citizenship_plot3
# This one seems too crowded to be an effective visualization
```

```{r}
citizen1 <- atuscps %>% 
  dplyr::select(tucaseid, region, state, race, citizen) %>%
  filter(citizen == "yes") %>%
  group_by(race, region) %>%
  summarize(
    is_citizen = n()
  )
citizen1
```

```{r}
citizen2 <- atuscps %>% 
  dplyr::select(tucaseid, region, state, race, citizen) %>%
  filter(citizen == "no") %>%
  group_by(race, region) %>%
  summarize(
    not_citizen = n()
  )
citizen2
```

```{r}
citizen3 <- citizen1 %>%
  left_join(citizen2, by = c("race", "region")) %>%
  mutate(citizen_ratio = is_citizen/(is_citizen + not_citizen))
citizen3
# Not sure how to proceed
```

## Country of Origin Across Races
```{r}
country_plot <- ggplot(data = atuscps, mapping = aes(x = race, fill = country_born)) +
  geom_bar(position = "fill")
country_plot
```

By Region:
```{r}
country_plot2 <- ggplot(data = atuscps, mapping = aes(x = race, fill = country_born)) +
  geom_bar(position = "fill") +
  facet_wrap(~ region)
country_plot2
```

By State:
```{r}
country_plot3 <- ggplot(data = atuscps, mapping = aes(x = race, fill = country_born)) +
  geom_bar(position = "fill") +
  facet_wrap(~ state) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
country_plot3
# Too crowded again
```

# Income Brackets
## Race
```{r}
race_income <- ggplot(data = atuscps, mapping = aes(x = famincome, fill = race)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
race_income
```

Regional Differences
```{r}
race_income_reg <- ggplot(data = atuscps, mapping = aes(x = famincome, fill = race)) +
  geom_bar(position = "fill") +
  facet_wrap(~ region)+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
race_income_reg
```

State Differences
```{r}
race_income_state <- ggplot(data = atuscps, mapping = aes(x = famincome, fill = race)) +
  geom_bar(position = "fill") +
  facet_wrap(~ state) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
race_income_state
# Too crowded
```

## Citizenship
```{r}
racecitizen_income <- ggplot(data = atuscps, mapping = aes(x = famincome, fill = race)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_wrap(~ citizen)
racecitizen_income
```

Regional
```{r}
racecitizen_income2 <- ggplot(data = atuscps, mapping = aes(x = famincome, fill = race)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(citizen ~ region)
racecitizen_income2
```

State (Can't visualize effectively)
```{r}
# racecitizen_income3 <- ggplot(data = atuscps, mapping = aes(x = famincome, fill = race)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  facet_grid(citizen ~ state)
# racecitizen_income3
# Cannot be visualized effectively
```

# Median Household Income

# Income Disparities (Top 1% vs. Bottom 99%)