---
title: "Group Project - STAT1261"
author: "Mahima Sindhu"
date: "11/22/2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyr)
library(dplyr)

library(atus)
data("atuscps")
tibble1 <- as_tibble(atuscps) %>%
  na.omit()
```

```{r}
library(ggplot2)
library(RColorBrewer)
tibble1 <- as_tibble(atuscps) %>%
  na.omit()

plotcps <- ggplot(data=tibble1) +
  geom_bar(mapping=aes(x=famincome, fill=home_type)) +
  xlab("Family Income") +
  ylab("Count") +
  scale_color_brewer("Set1")

plotcps + theme(axis.text.x = element_text(angle = 27, size = 6)) 
```

```{r}
library(ggplot2)
library(RColorBrewer)

plotdps <- ggplot(data=tibble1) +
  geom_bar(mapping=aes(x=home_type, fill=famincome)) + 
  xlab("Home Type") +
  ylab("Count") +
  scale_color_brewer("Set1")

plotdps + theme(axis.text.x = element_text(angle = 27, size = 6))
```

VISUALIZATION ANALYSIS:
Color-coded simple stacked bar charts, using ggplot2, were chosen to visualize the data because this method is useful in comparing the total amounts across each segmented group. Additionally, stacked bar charts have the advantages of being able to easily view the data and identify potential outliers.

When considering the relationship between the type of home (own, rent, occupy without rent) and family income, the two stacked bar charts visualize the following. No matter what family income range one falls under, there seems to be worth and demand in owning a house. This means even within a humble, lower family income range, there are still home-owners present. This offers the potentially interesting finding that people of lower socioeconomic classes are still trying to chase “the American Dream” or stability, despite renting being a more feasible option for their financial situation. It must be said of course, that this idea is not generalizable to that socioeconomic class as a whole because as both stacked bar charts suggest visually, for lower socioeconomic classes, the renter population still remains more popular than the homeowner population. On the contrary, we see that in the middle to higher income brackets, there are certainly a much larger count of homeowners compared to its opposing category, renters. 

```{r}
library(atus)
chart = table(atuscps$famincome, atuscps$home_type) 
chart

chisq.test(chart)
```

RELATIONSHIP SIGNIFICANCE ANALYSIS (Pearson's Chi-Squared Test):
As the p-value (<2.2e-16) is less than the .05 significance level, and the chi-squared value is notably large (26953), we reject the null hypothesis and conclude that the two variables, family income and home type (own, rent, occupy without rent), are in fact dependent and have a significant relationship. 

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(atus)

atus2 <- cbind(atuscps, atusresp) 
atusresp$hh_size <- as.character(as.numeric(atusresp$hh_size))

atus3 <- atus2 %>%
  select(famincome, hh_size)

atus4 <- as_tibble(atus3)

chart1 <- ggplot(data=atus4) +
  geom_bar(mapping=aes(x=famincome, fill=hh_size)) + 
  xlab("Family Income") +
  ylab("Count") +
  scale_color_brewer("Set1") 

chart1 + theme(axis.text.x = element_text(angle = 27, size = 6))
```

```{r}
library(tidyr)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
library(atus)

atus2 <- cbind(atuscps, atusresp) 
atusresp$hh_size <- as.character(as.numeric(atusresp$hh_size))

atus3 <- atus2 %>%
  select(famincome, hh_size)

atus4 <- as_tibble(atus3)

chart14 <- ggplot(data=atus4) +
  geom_bar(mapping=aes(x=hh_size, fill=famincome)) + 
  xlab("Household Size") +
  ylab("Count") +
  scale_color_brewer("Set1") 

chart14 + theme(axis.text.x = element_text(angle = 27, size = 6))
```

VISUALIZATION ANALYSIS:
The stacked bar charts depicting the relationship between household size (1-16) and family income suggest that there is a varied and potentially similar distribution of household size across each family income range. Further, in taking a closer look at the distribution of family income across household sizes, it is clear that there is a varied distribution, with no income group particularly prominent in each household size. Hence, we must use statistical analyses to see if household size and family income have a significant relationship.

```{r}
library(atus)
chart2 = table(atuscps$famincome, atusresp$hh_size) 
chart2

chisq.test(chart2)
```

RELATIONSHIP SIGNFICANCE ANALYSIS (Pearson's Chi-Squared Test):
As the p-value (<2.2e-16) is less than the .05 significance level, and the chi-squared value is notably large (21089), we reject the null hypothesis and conclude that the two variables, family income and household size (1-16), are in fact dependent and have a significant relationship. 

