---
title: "Final Project Proposal"
subtitle: Group 11
date: "11/8/2020"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Group Members
* James Brennan
* Linh Huynh
* Zachary Palmer
* Mahima Sindhu

### Refined Goals of Analysis
  Our group is using the `atus` database to investigate whether various 
  demographic variables correlate to family income. Specifically, we will be
  looking at region, state, educational attainment, home ownership (rent vs. own),
  household size, race, and time usage. After visualizing the data, we will test the 
  correlations between each of these demographic variables and income using
  paired t-tests or other related statistical analyses. Using the results from
  these statistical analyses, we will determine relationships between the
  demographic variables and how they collectively impact family income and how
  income potentially affects time usage. This information could be useful to
  better inform policy regarding issues relevant to specific demographics.

### Preliminary Data Exploration

#### Zach
```{r, warning=FALSE, include=FALSE}
library(atus)
library(tidyquant)
library(dplyr)
library(tidyr)
library(ggplot2)
library(scales)
```

```{r}
time_data = ungroup(atusact)
time_data2 = time_data %>% 
  group_by(tucaseid, tiercode) %>%
  summarise(tier1_2Code = tiercode %/% 100,
            tier1 = tiercode %/% 10000,
            tier2 = (tiercode %% 10000) %/% 100, 
            tier3 = (tiercode %% 10000) %% 100,
            dur = dur) %>%
  inner_join(atuscps) %>% 
  select(tiercode, tier1, dur, region, state, famincome) %>%
  drop_na()
time_data2 %>% head(5)
```

```{r}
time_data2$tier1 = as.factor(time_data2$tier1)
ggplot(data = time_data2, mapping = aes(x = famincome, y = dur, fill = tier1)) + 
  geom_col(position = "fill") + 
  coord_flip()
```

#### James
```{r}
# Find average family income by educational attainment
avg_income_by_edu <- atuscps 
avg_income_by_edu <- avg_income_by_edu %>%
  separate(famincome, into=c('income_low','income_high'), sep='-',convert=TRUE)%>%
  mutate(income_low = as.integer(income_low)) %>%
  drop_na() %>%
  mutate(fam_income_mid = (income_high+income_low)/2) %>%
  group_by(education=edu) %>%
  summarise(N = n(),
            avg_income = mean(fam_income_mid)) %>%
  arrange(avg_income)
avg_income_by_edu %>% head(5)
```
```{r}
#Visualize results
ggplot(data=avg_income_by_edu, aes(x=education, y=avg_income)) +
  geom_bar(stat='identity') +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
# Investigate by region
# Find average family income by educational attainment
avg_income_by_edu_region <- atuscps 
avg_income_by_edu_region <-avg_income_by_edu_region %>%
  separate(famincome, into=c('income_low','income_high'), sep='-',convert=TRUE)%>%
  mutate(income_low = as.integer(income_low)) %>%
  drop_na() %>%
  mutate(fam_income_mid = (income_high+income_low)/2) %>%
  group_by(region,edu) %>%
  summarise(N = n(),
            avg_income = mean(fam_income_mid)) %>%
  arrange(avg_income)
avg_income_by_edu_region %>% head(5)
```

```{r}
# Compare regions by income
region_df <- avg_income_by_edu_region %>%
  group_by(region) %>%
  summarise(N=n(),
            avg_income = mean(avg_income)) %>%
  arrange(avg_income)
ggplot(region_df, aes(x=region, y=avg_income, label=avg_income)) +
  geom_bar(stat='identity') +
  geom_text(aes(label = signif(avg_income, digits = 3)))
```

```{r}
# Compare income by region and edu
ggplot(avg_income_by_edu_region, aes(x=region, y=avg_income)) +
    geom_bar(position='stack',stat='identity') +
    scale_y_continuous(labels = function(x) format(x, scientific = FALSE)) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    geom_text(aes(label = signif(avg_income, digits = 3)), angle = 90, 
              nudge_y = -10000, color = 'white') +
    facet_grid(~ edu)
```

#### Mahima
```{r}
data("atuscps")
cps <- as_tibble(atuscps) 
cps
```


```{r}
library(ggplot2)
plotcps <- ggplot(data=atuscps) +
  geom_bar(mapping=aes(x=famincome, fill=home_type)) + 
  xlab("Family Income") +
  ylab("Count") 
plotcps + theme(axis.text.x = element_text(size = 3))
```

#### Linh
Preparing Data Set
```{r}
reg_income <- atuscps %>%
  select(region, famincome) %>%
  separate(famincome, into = c("low_famincome", "high_famincome"), sep = "-") %>%
  mutate(
    low_famincome = as.integer(low_famincome), 
    high_famincome = as.integer(high_famincome)
  ) %>%
  filter(
    !low_famincome %in% NA,
    !high_famincome %in% NA
  )
reg_income %>% head(5)
```

Finding Original Income Brackets
```{r}
levels(atuscps$famincome)
```

Bounds of Family Income
```{r}
bounds_reg_income <- reg_income %>%
  group_by(region) %>%
  summarise(
    mean_low_fincome = mean(low_famincome),
    mean_high_fincome = mean(high_famincome)
  )
bounds_reg_income %>% head(5)
```

New Income Brackets
```{r}
i_brackets <- bounds_reg_income %>%
  mutate(
    mean_low_fincome = round(mean_low_fincome),
    mean_high_fincome = round(mean_high_fincome)
  ) %>%
  unite(income_bracket, mean_low_fincome, mean_high_fincome, sep = "-")
i_brackets %>% head(5)
```

Racial Breakdown
```{r}
num_resp <- nrow(atuscps); num_resp
```

```{r}
race <- atuscps %>%
  select(race) %>%
  group_by(race) %>%
  summarise(
    num_race = n()
  ) %>%
  mutate(percent_race = percent((num_race/num_resp))) 
race %>% head(5)
```

```{r}
race_plot <- ggplot(data = atuscps, aes(x = factor(1), fill = race)) +
  geom_bar(width = 1) +
  coord_polar(theta = "y")
race_plot 
```

Citizenship Status Across Races
```{r}
citizenship_plot <- ggplot(data = atuscps, mapping = aes(x = race, fill = citizen)) +
  geom_bar(position = "fill")
citizenship_plot
```

Country of Origin Across Races
```{r}
country_plot <- ggplot(data = atuscps, mapping = aes(x = race, fill = country_born)) +
  geom_bar(position = "fill")
country_plot 
```

Family Income Based on Race
```{r}
race_income <- ggplot(data = atuscps, mapping = aes(x = famincome, fill = race)) +
  geom_bar(position = "fill") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
race_income
```